Estimating demographic bias on tests of children's early vocabulary

### 2022-01-16 paper/cogsci.Rmd
color-blind-friendly colors

@@

# should we report t-tests? 
table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
```

@@

```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs. age by demographic group, from the Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
  xlab("Age (months)") + ylab("Language Ability")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "SES")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic() + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "ethnicity")

@@

```{r demo-ability}
# show density plots of child ability based on group models
# sex
sex <- d_demo %>% filter(!is.na(sex), production!=0) %>%
  mutate(theta_sex = fscores(mod_intuitive_sex)) 
# cor.test(eth$production, eth$theta_eth) # .98

sex_m <- sex %>% group_by(sex) %>%
  summarise(ability = mean(theta_sex, na.rm=T))

p1 <- sex %>%
  ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=sex_m, aes(xintercept=ability, color=sex))

# SES
ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
  mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
  rename(SES=ses_group)
# cor.test(eth$production, eth$theta_eth) # .98

ses_m <- ses %>% group_by(SES) %>%
  summarise(ability = mean(theta_ses, na.rm=T))

p2 <- ses %>%
  ggplot(aes(x=theta_ses, color=SES, fill=SES)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=ses_m, aes(xintercept=ability, color=SES))

# ethnicity
eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
  rename(`Ethnic Group`=eth_group)
# cor.test(eth$production, eth$theta_eth) # .98

eth_m <- eth %>% group_by(`Ethnic Group`) %>%
  summarise(ability = mean(theta_eth, na.rm=T))

p3 <- eth %>%
  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`))

ggarrange(p1, p2, p3)
```

>>>>

```{r colors}
# color-blind palette
cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
sex_colors = cbp1[2:3]
eth_colors = cbp1[4:5]
ses_colors = cbp1[6:7]
```

@@

# should we report t-tests? 
#table(subset(d_demo, age>22 & age<26)$sex) # 395 female, 408 male
#table(subset(d_demo, age>22 & age<26)$ses_group) # 1038 high-SES, 140 low-SES
#table(subset(d_demo, age>22 & age<26)$eth_group) # 144 nonwhite, 603 white
```

@@

```{r baseline-ability-vs-age, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=6, set.cap.width=T, num.cols.cap=1, fig.cap = "Language ability vs. age by demographic group, from the Rasch model."}
p1 <- d_demo %>% filter(!is.na(sex)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=sex)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(type=sex_colors)
  #theme(legend.position = "top")

p2 <- d_demo %>% filter(!is.na(ses_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=ses_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") +
  scale_color_discrete(name = "SES", type=ses_colors) #+ 
  #theme(legend.position = "top")

p3 <- d_demo %>% filter(!is.na(eth_group)) %>%
  ggplot(aes(x=jitter(age), y=ability, color=eth_group)) + 
  geom_point(alpha=.1) + geom_smooth() + theme_classic(base_size=9) + 
  xlab("Age (months)") + ylab("Language Ability") + 
  scale_color_discrete(name = "ethnicity", type=eth_colors) #+ 
  #theme(legend.position = "top")

@@

```{r demo-ability, out.width="\\linewidth", fig.env = "figure", fig.pos = "H", fig.align='center', fig.width=3.3, fig.height=5.5, set.cap.width=T, num.cols.cap=1, fig.cap = "Density plots of language ability for children aged 23-25 months, based on the group models."}
# show density plots of child ability based on group models
# sex
sex <- d_demo %>% filter(!is.na(sex), production!=0) %>%
  mutate(theta_sex = fscores(mod_intuitive_sex)) 
# cor.test(eth$production, eth$theta_eth) # .98

sex_m <- sex %>% group_by(sex) %>%
  filter(age<26, age>22) %>%
  summarise(ability = mean(theta_sex, na.rm=T))

p1 <- sex %>% filter(age<26, age>22) %>%
  ggplot(aes(x=theta_sex, color=sex, fill=sex)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=sex_m, aes(xintercept=ability, color=sex)) +
    scale_color_discrete(type=sex_colors) + 
    scale_fill_discrete(type=sex_colors)

# SES
ses <- d_demo %>% filter(!is.na(ses_group), production!=0) %>%
  mutate(theta_ses = fscores(mod_intuitive_ses)) %>%
  filter(age<26, age>22) %>%
  rename(SES=ses_group)
# cor.test(eth$production, eth$theta_eth) # .98

ses_m <- ses %>% group_by(SES) %>%
  filter(age<26, age>22) %>%
  summarise(ability = mean(theta_ses, na.rm=T))

p2 <- ses %>%
  ggplot(aes(x=theta_ses, color=SES, fill=SES)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=ses_m, aes(xintercept=ability, color=SES)) + 
    scale_color_discrete(type=ses_colors) +
    scale_fill_discrete(type=ses_colors)

# ethnicity
eth <- d_demo %>% filter(!is.na(eth_group), production!=0) %>%
  mutate(theta_eth = fscores(mod_intuitive_eth)) %>%
  filter(age<26, age>22) %>%
  rename(`Ethnic Group`=eth_group)
# cor.test(eth$production, eth$theta_eth) # .98

eth_m <- eth %>% group_by(`Ethnic Group`) %>%
  filter(age<26, age>22) %>%
  summarise(ability = mean(theta_eth, na.rm=T))

p3 <- eth %>%
  ggplot(aes(x=theta_eth, color=`Ethnic Group`, fill=`Ethnic Group`)) +
    geom_density(alpha=.2) + theme_classic() + xlab("Ability") + 
    geom_vline(data=eth_m, aes(xintercept=ability, color=`Ethnic Group`)) + 
    scale_color_discrete(type=eth_colors) + 
    scale_fill_discrete(type=eth_colors)
